OpenMPIR

### 2017-09-08 paper.tex
citations and modified example

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} which allow such transformations
to be written once with relative ease and apply to a variety of source
front-end languages such as C or C++.  Historically, the IR of mainstream
compilers such as LLVM~\cite{llvm} or GCC's Gimple~\cite{} haven't suppored
parallel constructs. As a consequence, compilers utilizing these IRs haven't
had the ability to reason about parallelism. The recent work of Schardl et
al.~\cite{tapir} has shown that the LLVM IR can be extended to represent
fork-join parallelism without requiring a major rewrite.  They did so by
extending the LLVM instruction set with three instructions capable of
representing fork-join parallelism. To benchmark performance and test accuracy,
they wrote a frontend which translated programs written with Cilk, a parallel
programming model for C/C++, into the Tapir IR. They also implemented a backend
to \textbf{lower} Tapir programs to vanilla LLVM IR with embedded Cilk runtime
calls. They suggested that a similar approach could be taken with OpenMP and
other such frameworks.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc, mpi}
\pmnote{I would add MPI and UPC/PGAS here too}. From libraries, to language
extensions, to standalone languages, to combinations of the above,
fragmentation of parallel programming models is a problem that is only getting
worse. This means that when writing tooling, optimizations, or analyses, one
has to choose which model to target as well as understand points of contention
and interoperability between them. A common difficulty among all of these is
that reasoning about parallel programs is hard and compilers can potentially
help reduce this burden. This is exactly analogous the problem LLVM solves:
compiler optimizations are hard, so sharing them is valuable. Tapir is a
demonstration that the the LLVM approach to compilation of serial code can be
effectively extended to help reason about parallel code.

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{tapir-example}
we show the input OpenMP program and a portion of the corresponding Tapir code 
that is generated.

\begin{figure*}
\begin{tabular}{| c c c | c c c |}
\hline
 & 
\begin{lstlisting}
int fib(int n) 
{                    // A
  if (n < 2)
    return n;
  else
  {
    int x, y;
    #pragma omp task
      x = fib(n-1);  // B
    #pragma omp task
      y = fib(n-2);  // C
    #pragma omp taskwait
    return (x+y);    // D
  }
}
\end{lstlisting}
& & &
\begin{lstlisting}[language=llvm]
if.end:                                 
  detach label %det.achd, label %det.cont

@@

The listed Tapir IR code corresponds to lines 7 through 12 of the C source
code.  Upon entering the \texttt{else} branch, the code immediately detaches
the basic block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
actually a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizataions turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

While we would like to distinguish between these, it is difficult to do so with
only performance counter information. A more thorough investigation would
require deep knowledge of each of the frontends and runtimes. While the
importance of optimizations of parallel code is clear, we hypothesize that most
of the disparity present in these results is a function of runtime
implementation differences. As justification for this hypothesis, consider
the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
and the code leaves little room for optimization, algorithmic changes
notwithstanding. In other words, it is effectively a benchmark of how fast a
runtime can execute fork-join parallelism. \gsnote{still worth looking into the
optimization I mention in the implemtnation section.}

@@

A seperate but related area in LLVM extensions is the area of using memory

>>>>

Compilers implement the majority of their analyses and optimizations on
intermediate representations \textbf{(IRs)} which allow such transformations
to be written once with relative ease and apply to a variety of source
front-end languages such as C or C++.  Historically, the IR of mainstream
compilers such as LLVM~\cite{lattner2004llvm} or GCC's
Gimple~\cite{merrill2003generic} haven't suppored parallel constructs. As a
consequence, compilers utilizing these IRs haven't had the ability to reason
about parallelism. The recent work of Schardl et al.~\cite{tapir} has shown
that the LLVM IR can be extended to represent fork-join parallelism without
requiring a major rewrite.  They did so by extending the LLVM instruction set
with three instructions capable of representing fork-join parallelism. To
benchmark performance and test accuracy, they wrote a frontend which translated
programs written with Cilk, a parallel programming model for C/C++, into the
Tapir IR. They also implemented a backend to \textbf{lower} Tapir programs to
vanilla LLVM IR with embedded Cilk runtime calls. They suggested that a similar
approach could be taken with OpenMP and other such frameworks.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi} . From libraries, to language extensions, to standalone languages, to
combinations of the above, fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one has to choose which model to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous the problem LLVM solves: compiler optimizations are hard, so sharing
them is valuable. Tapir is a demonstration that the the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
parallel fragment of the \texttt{fib} example from Section~\ref{Sec:OpenMP}. In
Figure~\ref{Fig:Example} we show what the input OpenMP code and the Tapir code
is generated.

\begin{figure*}
\begin{tabular}{| c c c | c c c |}
\hline
 & 
\begin{lstlisting}
  #pragma omp task
    x = fib(n-1);  
  #pragma omp task
    y = fib(n-2);  
  #pragma omp taskwait
\end{lstlisting}
& & &
\begin{lstlisting}[language=llvm]
if.end:                                 
  detach label %det.achd, label %det.cont

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
actually a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizataions turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

While we would like to distinguish between these, it is difficult to do so with
only performance counter information. A more thorough investigation would
require deep knowledge of each of the frontends and runtimes. While the
importance of optimizations of parallel code is clear, we hypothesize that most
of the disparity present in these results is a function of runtime
implementation differences. As justification for this hypothesis, consider
the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
and the code leaves little room for optimization (algorithmic rewrites
notwithstanding). In other words, it is effectively a benchmark of how fast a
runtime can execute fork-join parallelism .

@@

One interesting area for future work would be combining Tapir with work on
reasoning about locality in LLVM. For example, Hayashi et al.
\cite{hayashi2015llvm} use address spaces in LLVM to reason about memory
locality in partitioned global address space (PGAS) systems. By combining
reasoning about code concurrency and locality, one could potentially apply
optimizations to better schedule and co-locate code and data.
